Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-11-06 evaluation.tex
edits

\begin{enumerate}
\item  $I_1$: An enrollment record should always refer to an existing student and an existing course.
\item  $I_2$: The capacity (\C{c\_capacity}) of a course should always be a
  non-negative quantity.
\end{enumerate}
\noindent Both $I_1$ and $I_2$ can be violated under weak isolation.
$I_1$ can be violated, for example, when \C{deregister\_student} runs
concurrently with \C{enroll}, both at RC isolation. While the former
transaction removes the student record after checking that no
enrollments for that student exists, the latter transaction
concurrently adds an enrollment record after checking the student
exists.  Both can succeed concurrently, resulting in an invalid state.
Invariant $I_2$ can be violated by two \C{enroll}s, both reading
\C{c\_capacity}=1, and both (atomically) decrementing it, resulting in
\C{c\_capacity}=-1.  We ran \thetool on the Courseware application
(Fig.~\ref{fig:courseware_code}) after annotating transactions with
their respective guarantees, and asserting $I = I_1 \wedge I_2$ as the
correctness condition. The guarantees $G_e$ and $G_d$ for \C{enroll}
and \C{deregister\_student} transactions, respectively, are shown below:
\begin{smathpar}
  \begin{array}{lcl}
    G_e(\stg,\stg') & \Leftrightarrow & \stg_s'=\stg_s
      \conj \exists\C{cid}.\exists\C{sid}.\\
    & & \hspace*{0.3in} \stg_c' = \stg_c \bind \lambda c.~
        \itel{c.\C{c\_id}=\C{cid}\\
    & & \hspace*{1.15in}}
          {\existsl(c',~c'.\C{id}=c.\C{id} \wedge
              c'.\C{c\_name}=c.\C{c\_name} \wedge c'.\C{c\_capacity}\ge0,
              ~\{c'\})\\
    & & \hspace*{1.15in}}
          {\{c\}}\\
    & & \hspace*{0.15in}\conj \stg_e = \stg_e' \bind \lambda e.~ 
        \itel{e.\C{e\_c\_id}=\C{cid} \wedge e.\C{e\_s\_id}=\C{sid}}
          {\emptyset}{\{e\}}\\
%   & & \hspace*{1.15in}} 
%         {\emptyset} {\{e\}}\\
    G_d(\stg,\stg') & \Leftrightarrow & \stg_c' = \stg_c \conj
      \stg_e' = \stg_e \conj \exists \C{sid}.~
      \itel{\forall(e \in \stg_e).~e.\C{e\_s\_id}\neq\C{sid}\\
%   & & \hspace*{0.3in} \itel{\forall(e \in
%           \stg_e).~e.\C{e\_s\_id}\neq\C{sid}\\
    & & \hspace*{1.55in}}
          {\stg_s' = \stg_s \bind \lambda s.~
           \itel{s.\C{id}=\C{sid}}{\emptyset}{\{s\}}\\
    & & \hspace*{1.55in}}
          {\stg_s' = \stg_s}\\
  \end{array}
\end{smathpar}
For the sake of this presentation we split $\stg$ into three disjoint
sets of records, $\stg_s$, $\stg_c$, and $\stg_e$, standing for
\C{Student}, \C{Course}, and \C{Enrollment} tables, respectively.
Observing that the set language $\SL$ (Sec.~\ref{sec:inference}),
besides being useful for automatic verification, also facilitates
succinct expression of transaction semantics, we define $G_e$ and
$G_d$ in a combination of FOL and $\SL$. $G_e$ essentially says that
the \C{enroll} transaction leaves the \C{Student} table unchanged,
while it may update the \C{c\_capacity} field of a \C{Course} record
to a non-negative value (even when it doesn't update, it is the case
that $c'.\C{c\_capacity}\ge0$, because $c'=c$, and $c\in\stg_c$, and
we know that $I_2(\stg_c)$). $G_e$ also conveys that \C{enroll} might
insert a new \C{Enrollment} record by stating that $\stg_e$, the
\C{Enrollment} table in pre-state, contains all records $e$ from
$\stg_e'$, the table in post-state, except when $e.\C{e\_c\_id}$ and
$e.\C{e\_s\_id}$ match \C{cid} and \C{sid}, respectively. The
guarantee $G_d$ of \C{deregister\_student} asserts that the
transaction doesn't write to \C{Course} and \C{Enrollment} tables. The
transaction might however delete a \C{Student} record bearing an
\C{id}=\C{sid} (formally, $\stg_s' = \stg_s \bind \lambda s.~
\itel{s.\C{id}=\C{sid}}{\emptyset}{\{s\}}$), for some \C{sid} for
which no corresponding \C{Enrollment} records are present in the
pre-state (in other words, $\forall(e \in
\stg_e).~e.\C{e\_s\_id}\neq\C{sid}$).

>>>>

\begin{enumerate}
\item  $I_1$: An enrollment record should always refer to an existing student and an existing course.
\item  $I_2$: The capacity (\C{c\_capacity}) of a course should always be a
  non-negative quantity.
\end{enumerate}
\noindent Both $I_1$ and $I_2$ can be violated under weak isolation.
$I_1$ can be violated, for example, when \C{deregister\_student} runs
concurrently with \C{enroll}, both at RC isolation. While the former
transaction removes the student record after checking that no
enrollments for that student exists, the latter transaction
concurrently adds an enrollment record after checking the student
exists.  Both can succeed concurrently, resulting in an invalid state.
Invariant $I_2$ can be violated by two \C{enroll}s, both reading
\C{c\_capacity}=1, and both (atomically) decrementing it, resulting in
\C{c\_capacity}=-1.  We ran \thetool on the Courseware application
(Fig.~\ref{fig:courseware_code}) after annotating transactions with
their respective guarantees, and asserting $I = I_1 \wedge I_2$ as the
correctness condition. The guarantees $G_e$ and $G_d$ for \C{enroll}
and \C{deregister\_student} transactions, respectively, are shown below:
\begin{smathpar}
  \begin{array}{lcl}
    G_e(\stg,\stg') & \Leftrightarrow & \stg_s'=\stg_s
      \conj \exists\C{cid}.\exists\C{sid}.\\
    & & \hspace*{0.3in} \stg_c' = \stg_c \bind \lambda c.~
        \itel{c.\C{c\_id}=\C{cid}\\
    & & \hspace*{1.15in}}
          {\existsl(c',~c'.\C{id}=c.\C{id} \wedge
              c'.\C{c\_name}=c.\C{c\_name} \wedge c'.\C{c\_capacity}\ge0,
              ~\{c'\})\\
    & & \hspace*{1.15in}}
          {\{c\}}\\
    & & \hspace*{0.15in}\conj \stg_e = \stg_e' \bind \lambda e.~ 
        \itel{e.\C{e\_c\_id}=\C{cid} \wedge e.\C{e\_s\_id}=\C{sid}}
          {\emptyset}{\{e\}}\\
%   & & \hspace*{1.15in}} 
%         {\emptyset} {\{e\}}\\
    G_d(\stg,\stg') & \Leftrightarrow & \stg_c' = \stg_c \conj
      \stg_e' = \stg_e \conj \exists \C{sid}.~
      \itel{\forall(e \in \stg_e).~e.\C{e\_s\_id}\neq\C{sid}\\
%   & & \hspace*{0.3in} \itel{\forall(e \in
%           \stg_e).~e.\C{e\_s\_id}\neq\C{sid}\\
    & & \hspace*{1.55in}}
          {\stg_s' = \stg_s \bind \lambda s.~
           \itel{s.\C{id}=\C{sid}}{\emptyset}{\{s\}}\\
    & & \hspace*{1.55in}}
          {\stg_s' = \stg_s}\\
  \end{array}
\end{smathpar}
For the sake of this presentation we split $\stg$ into three disjoint
sets of records, $\stg_s$, $\stg_c$, and $\stg_e$, standing for
\C{Student}, \C{Course}, and \C{Enrollment} tables, respectively.
Observing that the set language $\SL$ (Sec.~\ref{sec:inference}),
besides being useful for automatic verification, also facilitates
succinct expression of transaction semantics, we define $G_e$ and
$G_d$ in a combination of FOL and $\SL$. $G_e$ essentially says that
the \C{enroll} transaction leaves the \C{Student} table unchanged,
while it may update the \C{c\_capacity} field of a \C{Course} record
to a non-negative value (even when it doesn't update, it is the case
that $c'.\C{c\_capacity}\ge0$, because $c'=c$, and $c\in\stg_c$, and
we know that $I_2(\stg_c)$). $G_e$ also conveys that \C{enroll} might
insert a new \C{Enrollment} record by stating that $\stg_e$, the
\C{Enrollment} table in the pre-state, contains all records $e$ from
$\stg_e'$, the table in the post-state, except when $e.\C{e\_c\_id}$ and
$e.\C{e\_s\_id}$ match \C{cid} and \C{sid}, respectively. The
guarantee $G_d$ of \C{deregister\_student} asserts that the
transaction doesn't write to \C{Course} and \C{Enrollment} tables. The
transaction might however delete a \C{Student} record bearing an
\C{id}=\C{sid} (formally, $\stg_s' = \stg_s \bind \lambda s.~
\itel{s.\C{id}=\C{sid}}{\emptyset}{\{s\}}$), for some \C{sid} for
which no corresponding \C{Enrollment} records are present in the
pre-state (in other words, $\forall(e \in
\stg_e).~e.\C{e\_s\_id}\neq\C{sid}$).

### 2017-11-06 implementation.tex
edits

At the top-level, \thetool runs a loop that picks an unverified
transaction and progressively strengthens its isolation level until it
passes verification. If the selected data store provides a
serializable isolation level, and if the program is sequentially
correct, then the verification is guaranteed to succeed. Within the
loop, the \thetool first computes the various rely relations needed for
verification ($R$, $\R_l$, and $\R_c$). It then traverses the AST of a
transaction, applying the inference rules to construct a state
transformer, check its stability, and weaken it ($\stabilize{\cdot}$)
if it is not stable. The result of traversing the transaction's AST is
therefore a state transformer ($\F$) that is stable w.r.t $\R_l$, which
is also stabilized against $\R_c$ (using $\stabilize{\cdot}$), and
then checked against the transaction's stated guarantee ($G$). If the
check passes, then the guarantee is verified to check if it preserves
the invariant $I$. The successful result from both the checks results
in the transaction being certified correct under the current choice of
its isolation level. Successful verification of all transactions
concludes the top-level execution, returning the inferred isolation
levels as its output.

@@

\textbf{Real-World Isolation Levels} The axiomatization of the
isolation levels presented in \S\ref{sec:isolation} leaves out
certain nuances of their implementations on real data stores, which
need to be taken into account for verification to be effective in
practice.  We consider these into account while linking \thetool with
store-specific semantics (isolation specifications etc). As an
example, consider how PostgreSQL implements an \C{UPDATE} operation.
\C{UPDATE} first selects the records that meet the search criteria
from the snapshot against which it is executing (the snapshot is
established at the beginning of the transaction if the isolation level
is SI, or at the beginning of the \C{UPDATE} statement if the
isolation level is RC). The selected records are then visited in the
actual database (if they still exist), write locks are obtained, and
the update is performed, provided that the record still meets
\C{UPDATE}'s search criteria. If the record no longer meets the
search criteria (due to a concurrent update) the record is excluded
from the update, and the write lock is immediately released.
Otherwise, the record remains locked until the transaction commits.

Clearly, this sequence of events is not atomic, unlike the assumption
made by our formal model.  The implementation admits interference
between the updates of individual records that meet the search
criteria.  Nonetheless, through a series of relatively straightforward
deductions, we can show that PostgreSQL's \C{UPDATE} is in fact
equivalent (in behavior) to a sequential composition of two atomic
operations $c_1;c_2$, where $c_1$ is effectively a \C{SELECT}
operation with the same search criteria as \C{UPDATE}, and $c_2$ is
a slight variation of the original \C{UPDATE} that updates a
record only if a record with the same id is present in the set of records
returned by the \C{SELECT}. This transformation is summarized below:
\begin{smathpar}
\begin{array}{lcl}
\updatee{(\lambda x. ~e_1)}{(\lambda x.~e_2)}
&
\longrightarrow
&
\lete{y}{\selecte{(\lambda x.~e_1})}
     {\updatee{(\lambda x.~e_1 \wedge x.\idf\in\dom(y))}
              {(\lambda x.~e_2})}\\
\end{array}
\end{smathpar}
The intuition behind this translation is the observation that all
interferences possible during the execution of the \C{UPDATE} can be
accommodated between the time the records are selected from the
snapshot, and the time they are actually updated.  \thetool performs this
translation if the selected store is PostgreSQL, allowing it to reason
about \C{UPDATE} operations in a way that is faithful to its semantics
on PostgreSQL. \thetool also admits similar compensatory logic for
certain combinations of isolation levels and operations on MySQL.

\textbf{Set functions} SQL's \C{SELECT} query admits projections of
record fields, and also application of auxiliary functions such as
\C{MAX} and \C{MIN}, e.g., \C{SELECT MAX(ol\_o\_id) FROM
Order\_line WHERE $\ldots$}. We admit such extensions as set functions
in our DSL (e.g., \C{project}, \C{max}, \C{min}), and axiomatize their
behavior. For instance:
\begin{smathpar}
\begin{array}{lcl}
  s_2 \;=\;\C{project}\,s_1\,(\lambda z.~e) & \Leftrightarrow &
  \forall y.~y\in s_2 \Leftrightarrow  \exists(x \in s_1).~y = [x/z]e\\
  x \;=\; \C{max}\,s & \Leftrightarrow & x \in s \conj \forall(y \in
  s).~y\le x\\
\end{array}
\end{smathpar}
There are however certain set functions whose behavior cannot be
completely axiomatized in FOL. These include \C{sum}, \C{count} etc.
For these, we admit imprecise axiomatizations expressed as lemmas
over these functions.

>>>>

At the top-level, \thetool runs a loop that picks an unverified
transaction and progressively strengthens its isolation level until it
passes verification. If the selected data store provides a
serializable isolation level, and if the program is sequentially
correct, then the verification is guaranteed to succeed. Within the
loop, \thetool first computes the various rely relations needed for
verification ($R$, $\R_l$, and $\R_c$). It then traverses the AST of a
transaction, applying the inference rules to construct a state
transformer, checks its stability, and weakens it ($\stabilize{\cdot}$)
if it is not stable. The result of traversing the transaction's AST is
therefore a state transformer ($\F$) that is stable w.r.t $\R_l$, which
is also stabilized against $\R_c$ (using $\stabilize{\cdot}$), and
then checked against the transaction's stated guarantee ($G$). If the
check passes, then the guarantee is verified to check if it preserves
the invariant $I$. The successful result from both checks results
in the transaction being certified correct under the current choice of
its isolation level. Successful verification of all transactions
concludes the top-level execution, returning the inferred isolation
levels as its output.

@@

\textbf{Real-World Isolation Levels} The axiomatization of the
isolation levels presented in \S\ref{sec:isolation} leaves out
certain nuances of their implementations on real data stores, which
need to be taken into account for verification to be effective in
practice.  We take these into account while linking \thetool with
store-specific semantics (isolation specifications, etc.). As an
example, consider how PostgreSQL implements an \C{UPDATE} operation.
\C{UPDATE} first selects the records that meet the search criteria
from the snapshot against which it is executing (the snapshot is
established at the beginning of the transaction if the isolation level
is SI, or at the beginning of the \C{UPDATE} statement if the
isolation level is RC). The selected records are then visited in the
actual database (if they still exist), write locks are obtained, and
the update is performed, provided that each matched record still meets
\C{UPDATE}'s search criteria. If a record no longer meets the
search criteria (due to a concurrent update), it is excluded
from the update, and the write lock is immediately released.
Otherwise, the record remains locked until the transaction commits.

Clearly, this sequence of events is not atomic, unlike the assumption
made by our formal model because  the implementation admits interference
between the updates of individual records that meet the search
criteria.  Nonetheless, through a series of relatively straightforward
deductions, we can show that PostgreSQL's \C{UPDATE} is in fact
equivalent (in behavior) to a sequential composition of two atomic
operations $c_1;c_2$, where $c_1$ is effectively a \C{SELECT}
operation with the same search criteria as \C{UPDATE}, and $c_2$ is
a slight variation of the original \C{UPDATE} that updates a
record only if a record with the same id is present in the set of records
returned by the \C{SELECT}. This transformation is summarized below:
\begin{smathpar}
\begin{array}{lcl}
\updatee{(\lambda x. ~e_1)}{(\lambda x.~e_2)}
&
\longrightarrow
&
\lete{y}{\selecte{(\lambda x.~e_1})}
     {\updatee{(\lambda x.~e_1 \wedge x.\idf\in\dom(y))}
              {(\lambda x.~e_2})}\\
\end{array}
\end{smathpar}
The intuition behind this translation is the observation that all
interferences possible during the execution of the \C{UPDATE} can be
accommodated between the time the records are selected from the
snapshot, and the time they are actually updated.  \thetool performs this
translation if the selected store is PostgreSQL, allowing it to reason
about \C{UPDATE} operations in a way that is faithful to its semantics
on PostgreSQL. \thetool also admits similar compensatory logic for
certain combinations of isolation levels and operations on MySQL.

\textbf{Set functions} SQL's \C{SELECT} query admits projections of
record fields, and also application of auxiliary functions such as
\C{MAX} and \C{MIN}, e.g., \C{SELECT MAX(ol\_o\_id) FROM
Order\_line WHERE $\ldots$}, etc. We admit such extensions as set functions
in our DSL (e.g., \C{project}, \C{max}, \C{min}), and axiomatize their
behavior. For instance:
\begin{smathpar}
\begin{array}{lcl}
  s_2 \;=\;\C{project}\,s_1\,(\lambda z.~e) & \Leftrightarrow &
  \forall y.~y\in s_2 \Leftrightarrow  \exists(x \in s_1).~y = [x/z]e\\
  x \;=\; \C{max}\,s & \Leftrightarrow & x \in s \conj \forall(y \in
  s).~y\le x\\
\end{array}
\end{smathpar}
There are however certain set functions whose behavior cannot be
completely axiomatized in FOL. These include \C{sum}, \C{count} etc.
For these, we admit imprecise axiomatizations expressed as lemmas
over these functions.

### 2017-11-06 inference.tex
edits

The fragment of FOL that admits formulas with prenex signatures of the
form $\forall^2\exists^*$ is called the G\"odel-K\'almar-Sch\"utte ({\sf
GKS}) fragment~\cite{gks}, which is known to be decidable. The
language of encoding, however, is a combination of {\sf GKS} with (a).
$\Prop^0$, the theory from which quantifier-free propositions
($\varphi$) that encode object language expressions are drawn, and
(b). $\Prop^1$, the theory from which invariants ($I$) are drawn. Thus,
the encoding of the subset of $\SL$ described above is decidable if the
combination of ${\sf GKS} + \Prop^0 + \Prop^1$ is decidable. We write
$\SL[\Prop^0,\Prop^1]$ to highlight the parameterization of $\SL$ on
$\Prop^0$ and $\Prop^1$.  The discussion in the previous paragraph
points to the existence of non-trivial subsets in $\SL[\Prop^0,\Prop^1]$ that are
decidable:
\begin{theorem}
  There exists an $\SL'[\Prop^0,\Prop^1] \subset \SL[\Prop^0,\Prop^1]$
  such that $\SL'$ is decidable if ${\sf GKS}+\Prop^0+\Prop^1$ is
  decidable.
\end{theorem}
One interesting example of such an $\SL'$ is the subset described above: $\SL$ with
bind expressions confined to one level of nesting. We denote this
subset as $\SL^1[\Prop^0,\Prop^1]$, for which we assert decidability:
\begin{corollary}
  $\SL^1[\Prop^0,\Prop^1]$ is decidable if ${\sf GKS}+\Prop^0+\Prop^1$ is
  decidable.
\end{corollary}
$\SL^1$ is a useful subset of $\SL$, for it corresponds to $\txnimp$
programs without nested \C{foreach} loops. Observe that the
\C{new\_order} transaction (Fig.~\ref{fig:new_order_code}) belongs to
this subset.  Indeed, $\SL^1$, while being a restricted version of
$\SL$, is nonetheless expressive enough to cover all the benchmarks we
considered in Sec.~\ref{sec:case-studies}.

>>>>

The fragment of FOL that admits formulas with prenex signatures of the
form $\forall^2\exists^*$ is called the G\"odel-K\'almar-Sch\"utte ({\sf
GKS}) fragment~\cite{gks}, which is known to be decidable. The
language of encoding, however, is a combination of {\sf GKS} with (a).
$\Prop^0$, the theory from which quantifier-free propositions
($\varphi$) that encode object language expressions are drawn, and
(b). $\Prop^1$, the theory from which invariants ($I$) are drawn. Thus,
the encoding of the subset of $\SL$ described above is decidable if the
combination of ${\sf GKS} + \Prop^0 + \Prop^1$ is decidable. We write
$\SL[\Prop^0,\Prop^1]$ to highlight the parameterization of $\SL$ on
$\Prop^0$ and $\Prop^1$.  The discussion in the previous paragraph
points to the existence of non-trivial subsets in $\SL[\Prop^0,\Prop^1]$ that are
decidable:
\begin{theorem}
  There exist $\SL'[\Prop^0,\Prop^1] \subset \SL[\Prop^0,\Prop^1]$
  such that $\SL'$ is decidable if ${\sf GKS}+\Prop^0+\Prop^1$ is
  decidable.
\end{theorem}
One interesting example of such an $\SL'$ is the subset described above: $\SL$ with
bind expressions confined to one level of nesting. We denote this
subset as $\SL^1[\Prop^0,\Prop^1]$, for which we assert decidability:
\begin{corollary}
  $\SL^1[\Prop^0,\Prop^1]$ is decidable if ${\sf GKS}+\Prop^0+\Prop^1$ is
  decidable.
\end{corollary}
$\SL^1$ is a useful subset of $\SL$, for it corresponds to $\txnimp$
programs without nested \C{foreach} loops. Observe that the
\C{new\_order} transaction (Fig.~\ref{fig:new_order_code}) belongs to
this subset.  Indeed, $\SL^1$, while being a restricted version of
$\SL$, is nonetheless expressive enough to cover all the benchmarks we
considered in Sec.~\ref{sec:case-studies}.
